---
title: "CaseStudy"
author: "Jakub Hus"
date: "2026-01-17"
output: html_document
---

# Load data and define column names
```{r}
dane <- read.table(
  "C:/Users/Dell/Desktop/Eksploracja Danych/Case Study/PM10.dat",
  skip = 8, # Skip 8 lines at the top of the file
  header = FALSE,
  sep = "\t", # Tab-seperated values
  dec = "."
)

colnames(dane) <- c("log_PM10", "log_cars", "temp_2m", "wind_speed",
                    "tempdiff_25_2m", "wind_dir", "hour", "dayno")
```
# Basic data inspection
```{r}
head(dane)
```
```{r}
str(dane)
```
```{r}
summary(dane)
```
# Outlier detection via boxplots
# Each variable is plotted seperately to identify potential outliers that may affect model performance
```{r}
boxplot(dane$log_PM10,
        main = "Boxplot log_PM10",
        ylab = "log_PM10",
        las = 2)
```

```{r}
boxplot(dane$log_cars,
        main = "Boxplot log_cars",
        ylab = "log_cars",
        las = 2)
```
```{r}
boxplot(dane$temp_2m,
        main = "Boxplot temp_2m",
        ylab = "temp_2m",
        las = 2)
```
```{r}
boxplot(dane$wind_speed,
        main = "Boxplot wind_speed",
        ylab = "wind_speed",
        las = 2)
```
```{r}
boxplot(dane$tempdiff_25_2m,
        main = "Boxplot tempdiff_25_2m",
        ylab = "tempdiff_25_2m",
        las = 2)
```
```{r}
boxplot(dane$wind_dir,
        main = "Boxplot wind_dir",
        ylab = "wind_dir",
        las = 2)
```
# Feature engineering: create target variable and time-of-day feature
```{r}
# PM10: binary target variable based on EU air quality stanard (50 µg/m³)
# Since log_PM10 = ln(PM10), the threshold is log(50)
# 'exceed' -> pollution above the norm, 'norm' -> within acceptable range

dane$PM10 <- cut(
  dane$log_PM10,
  breaks = c(-Inf, log(50), Inf),
  labels = c("norm", "exceed"),
  right = FALSE
)

# day_part: categorises the hour of measurement into four time-of-day periods
# This captures daily traffic and weather patterns relevant to PM10 levels

dane$day_part <- cut(
  dane$hour,
  breaks = c(-Inf, 5, 11, 17, Inf),
  labels = c("night", "morning", "afternoon", "evening"),
  right = TRUE
)

table(dane$day_part)

```

```{r}
head(dane)
```
# Class distribution of the target variable
```{r}
table(dane$PM10)
prop.table(table(dane$PM10))
```
# Train, validation and test split
# 4 observations are held out as a demonstration prediction set (pred_set)
# The remaining data is split 75/25 into training and validation sets with stratification on PM10 to preserve class balance across splits
```{r}
library(caret)
set.seed(42)

id_pred <- sample(1:nrow(dane), 4)
pred_set <- dane[id_pred, ]
rest <- dane[-id_pred, ]

id_train <- createDataPartition(rest$PM10, p = 0.75, list = FALSE)
train_set <- rest[id_train, ]
valid_set <- rest[-id_train, ]

table(train_set$PM10)
table(valid_set$PM10)
```
# Random Forest model training
# Class weights are set to 1:3 (norm:exceed) to penalise missed exceedances more heavily than false alarms, detecting poor quality is the priority
```{r}
library(randomForest)

set.seed(42)
rf_model <- randomForest(
  PM10 ~ log_cars + temp_2m + wind_speed + tempdiff_25_2m + wind_dir + day_part,
  data = train_set,
  ntree = 500,
  importance = TRUE,
  classwt = c(norm = 1, exceed = 3)
)

rf_model
varImpPlot(rf_model)
```
# Validation: Random Forest predictions with custom threshold
# Default threshold of 0.5 is lowered to 0.23 to increase Recall for 'exceed'
```{r}
library(ggplot2)

pred_valid_rf <- predict(rf_model, newdata = valid_set, type = "prob")[, "exceed"]

pred_valid_rf_thr <- ifelse(pred_valid_rf >= 0.23, "exceed", "norm")
pred_valid_rf_thr <- factor(pred_valid_rf_thr, levels = c("norm", "exceed"))


cm <- confusionMatrix(
  pred_valid_rf_thr,
  valid_set$PM10,
  positive = "exceed"
)
# Visualise confusion matrix as a heatmap
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predykcja", "Rzeczywiste", "Liczba")


ggplot(cm_df, aes(x = Rzeczywiste, y = Predykcja, fill = Liczba)) +
  geom_tile(color = "black") +
  geom_text(aes(label = Liczba), size = 5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix – Random Forest",
    x = "Real",
    y = "Predicted"
  ) +
  theme_minimal()
```
# Model evaluation metrics - Random Forest
```{r}
cm$overall # Accuracy, Kappa
```
```{r}
cm$byClass # Sensitivity (Recall), Specificity, Precision, F1, etc.
```
# Prediction on 4-observation demonstration set - Random Forest
```{r}
pred_prob_rf <- predict(
  rf_model,
  newdata = pred_set,
  type = "prob"
)[, "exceed"]

threshold <- 0.23
pred_class_rf <- ifelse(pred_prob_rf >= threshold, "exceed", "norm")
pred_class_rf <- factor(pred_class_rf, levels = c("norm", "exceed"))

predykcje_rf <- data.frame(
  PM10_true = pred_set$PM10,
  P_exceed = round(pred_prob_rf, 3),
  PM10_pred = pred_class_rf
)

predykcje_rf
```
# CART decision tree model training
# An asymmetric loss matrix penalises false negatives 3x more than false positives, consistent with the cost assumption used in Random Forest
```{r}
library(rpart)
library(rpart.plot)
library(caret)

set.seed(42)

cart_model <- rpart(
  PM10 ~ log_cars + temp_2m + wind_speed + tempdiff_25_2m + wind_dir + day_part,
  data = train_set,
  method = "class",
  parms = list(
    loss = matrix(c(0, 1,
                    3, 0), nrow = 2)
  )
)

cart_model
```
# Decision tree visualisation
```{r}
rpart.plot(cart_model)
```
# Validation: CART predictions with custom threshold
```{r}
pred_valid_cart_prob <- predict(cart_model, newdata = valid_set, type = "prob")[, "exceed"]

pred_valid_cart_thr <- ifelse(pred_valid_cart_prob >= 0.23, "exceed", "norm")
pred_valid_cart_thr <- factor(pred_valid_cart_thr, levels = c("norm", "exceed"))

cm_cart_thr <- confusionMatrix(
  pred_valid_cart_thr,
  valid_set$PM10,
  positive = "exceed"
)
```

# Confusion matrix - CART
```{r}
cm_cart_df <- as.data.frame(cm_cart_thr$table)
colnames(cm_cart_df) <- c("Predykcja", "Rzeczywiste", "Liczba")

ggplot(cm_cart_df, aes(x = Rzeczywiste, y = Predykcja, fill = Liczba)) +
  geom_tile(color = "black") +
  geom_text(aes(label = Liczba), size = 5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix – CART",
    x = "Real",
    y = "Prediction"
  ) +
  theme_minimal()
```
# Model evaluation metrics - CART
```{r}
cm_cart_thr$overall
```
```{r}
cm_cart_thr$byClass
```
# Prediction on 4-observation demonstration set - CART
```{r}
pred_prob_cart <- predict(
  cart_model,
  newdata = pred_set,
  type = "prob"
)[, "exceed"]

threshold <- 0.23
pred_class_cart <- ifelse(pred_prob_cart >= threshold, "exceed", "norm")
pred_class_cart <- factor(pred_class_cart, levels = c("norm", "exceed"))

predykcje_cart <- data.frame(
  PM10_true = pred_set$PM10,
  P_exceed = round(pred_prob_cart, 3),
  PM10_pred = pred_class_cart
)

predykcje_cart
```






































































































































































